### 2013-01-04 ISMB/methods_ISMB.tex
stuff

\subsection{Probabilistic model}
Let $\Omega$ be an gap-free RNA alignment sequence, $S$ its associated secondary structure, 
then any sequence $s$ is assigned a probability proportional to its Boltzmann factor
\begin{align*}
  \mathcal{B}(s) &= e^\frac{-\PE{s}}{RT}, &&\text{with}&\PE{s}&:=\alpha\cdot\ES(s,S)+(1-\alpha)\cdot\EI(s,S,\Omega),
\end{align*}
where $R$ is the Boltzmann constant, $T$ the temperature in Kelvin, $\ES(s)$ and $\EI(s,S,\Omega)$ 
are the free-energy and isostericity contributions respectively (further described below), and $\alpha\in[0,1]$ is an arbitrary parameter that sets the relative weight for both contributions.

\subsubsection{Energy contribution}
The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, taking values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
\begin{align*}
  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
\end{align*}
where $\ES^{\beta}_{ab\to a'b'}$ is set to $0$ if $ab=\varnothing$ (no base-pair to stack onto), the tabulated free-energy of stacking pairs $(ab)/(a'b')$ in the Turner model if available, or $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. not in $\{\Gb\Ub,\Ub\Gb,\Cb\Gb,\Gb\Cb, \Ab\Ub\text{ or }\Ub\Ab\}$). This latter parameter allows to choose whether to simply penalize invalid base pairs, or forbid them altogether ($\beta = \infty$).
The loss of precision due to this simplification of the Turner model remains reasonable since the targeted secondary structure is fixed 
(e.g. multiloops do not account for base-specific contributions). Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
\subsubsection{Isostericity contribution}
The concept of isostericity score~\cite{Stombaugh2009} is based on the geometric discrepancy (superimposability) of two base-pairs, using individual additive contributions computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gap-free RNA alignment $\Omega$,  we define the contribution of the isostericity to the pseudo-energy as
\begin{align*}
  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
	\frac{
		\sum_{s'\in\Omega}
			\text{\ISO}((s_i',s_j'),(a,b))}
%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
{		
		|\Omega|
	}
\end{align*}
is the average isostericity of a base-pair in the candidate sequence, compared with the reference alignment.
The $\ISO$ function uses the {Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, with $0$ being assigned to a perfect isostericity, and a penalty of $10$ is used for missing entries.
The isostericity contribution will favor exponentially sequences that are likely to adopt a similar local conformation as the sequences contained in the alignment.

\subsection{Mutational profile of sequences}

Let $s$ be an RNA sequence, $S$ a reference structure, and $m\geq 0$ a desired number of mutations. 
We are interested in  the probability that a given position contains a specific nucleotide, over all sequences having at most $m$ mutations from $s$
%  under the SCFG derivation $S$ 
(formally $\mathbb{P}(s_i = x\mid s,\Omega, S,m)$). 
We define a variant of the
 \emph{Inside-Outside algorithm}~\cite{Lari1990}, allowing us to obtain the
 desired probability,  the two  functions
$\mathcal{Z}_*^*$ and $\mathcal{Y}_*^*$.

The former, defined in Equations~\eqref{eq:Z_in} and~\eqref{eq:Z_rec}, is analogous to the \emph{inside} algorithm.
It is the partition function, i.e. the sum of Boltzmann factors, over all sequences within $[i,j]$, 
knowing that position $i-1$ is composed of nucleotide $a$ (resp. $j+1$ is $b$), within 
$m$ mutations of $s$. The latter, defined by Equations~\eqref{eq:Y_in} and~\eqref{eq:Y_rec},
 computes the \emph{outside} algorithm, i.e.  
the partition function over sequences within $m$ mutations of $s$, restricted to two intervals $[0,i]\cup[j,n-1]$, and 
knowing  that position $i+1$ is composed of $a$ (resp. $j-1$ is $b$). A suitable combination of these terms, given in Equation~\eqref{eq:combine}, gives the total weight, and in turn the probability, of seeing a specific base at a given position.

%Drawing a parallel with stochastic context-free grammars (SCFG), for which the inside-outside algorithm was introduced, the set of sequences can be seen as the language of words having length $n$, generated from a stochastic context-free grammar.
%These rules would limit the number of mutations (i.e. )
% $S$ can be considered as constraining weighing on the shape of parse trees the derivations in a classic 
% of the {SCFG} generating  all  secondary structures of length  $|s|$.

\subsubsection{Definitions}
Let $B:=\left\{\Ab,\Cb,\Gb,\Ub\right\}$ be the set of nucleotides.
Given $s\in B^n$ an RNA sequence, let $s_i$ be the nucleotide at position $i$. Let $\Omega$ be a set of un-gapped RNA sequences of
length $n$, and $S$ a secondary structure without pseudoknots. 
Formally, if $(i,j)$ and $(k,l)$ are base pairs in $S$, there is no overlapping extremities
 $\{i,j\}\cap \{k,l\}=\varnothing$ and either the intersection is empty 
 ($[i,j]\cap[k,l]=\varnothing$) or one is included in the other ($[k,l]\subset[i,j]$ or 
 $[i,j]\subset[k,l])$.

Let us then remind a Hamming distance function $\delta: B^*\times B^* \to \mathbb{N}^+$, which takes two sequences $s'$ and $s''$ as input, $|s'|=|s''|$, and returns the number of differing positions.
Finally, let us denote by $E_{(i,j),ab \to ab'}^{\Omega,\beta}$ the local contribution of a base-pair $(i,j)$ to the pseudo-energy, such that
\begin{equation}
  E_{(i,j),ab \to a'b'}^{\Omega,\beta}  = \alpha \cdot\ES^\beta_{a b \to a' b'}+(1-\alpha)\cdot\EI^{\Omega}_{(i,j),a'b'}.
\end{equation}

\subsubsection{Inside}
\begin{figure}
\resizebox{\textwidth}{!}{\input{FigDPInside.pgf}}
\caption{Principle of the inside computation (aka partition function). Any sequence with $m$ mutations over an interval $[i,j]$ 
can be decomposed as a sequence over $[i+1,j]$ preceded by a, possibly mutated, base at $i$ 
(Unpaired case), a sequence over $[i+1,j-1]$ surrounded by some base-pair (Stacking-pair case), 
or as two sequences over $[i+1,k-1]$ and $[k+1,j]$, completed by some base-pair (General base-pairing case). 
In each case, one has to investigate any possible ways to distribute mutations between the different sequences 
and locally instanciated bases.}
\end{figure}

The \emph{Inside} function $\Z{i,j}{m}{a,b}$ is the partition function, i.e. the sum of Boltzmann factors, over all sequences in the interval $[i,j]$, at distance $m$ of $s_{[i,j]}$, and having flanking nucleotides $a$ and $b$ (at positions $i-1$ and $j+1$ respectively). 
Such terms can be defined by recurrence, for which the following initial conditions holds:

\begin{equation}
	\forall i \in [0,n-1]:\, \Z{i+1,i}{m}{a,b}=\left\{
	\begin{array}{ll}
		1 &\text{If } m = 0\\
		0 &\text{Otherwise.}
	\end{array}\right.
\label{eq:Z_in}
\end{equation}
In other words, the set of sequences at distance $m$ of the empty sequence is either empty if $m>0$, or restricted to the empty sequence, having energy $0$, if $m=0$. Since the energetic terms only depend on base pairs, they are not involved in the initial conditions.

@@

\subsubsection{Outside}	
\begin{figure}
\resizebox{\textwidth}{!}{\input{FigDPOutside.pgf}}
\caption{Principle of the outside computation. Note that the outside algorithm uses intermediate results from the inside algorithm, 
therefore its efficient implementation requires an implementation of the inside computation.}
\end{figure}
The \emph{Outside} function, $\mathcal Y$, is the partition function considering only the 
contributions of subsequences $[0,i]\cup[j,n-1]$ over the mutants of $s$ having exactly $m$ mutations between $[0,i]\cup[j,n-1]$ and whose nucleotide at position $i+1$ is $a$ 
(resp. in position $j-1$ it is $b$).
We define function $\Y{i,j}{m}{a,b}$ as a recurrence, and will use as initial conditions:
\begin{equation}
	\Y{-1,j}{m}{X,X}:=
		\displaystyle
	  \Z{j,n-1}{m}{X,X}
\label{eq:Y_in}
\end{equation}
The recurrence, as shown below, will increase the interval $[i,j]$ by decreasing $i$ when
it is not base paired. If it is with a position $k>j$, we increase $j$ to include it.
 Thus, when we need
to evaluate an interval as $(-1,j)$, all stems between $(0,j)$ are taken into account and the
structure between $(j,n-1)$ must be a set of independent stems. Therefore,
 all the outside energy between $[j,n-1]$ is
equal to $\Z{j,n-1}{m}{X,X}$, for any $X\in B$. The recursion itself is as follows.
\begin{equation}
	\Y{i,j}{m}{a,b} = \left\{
  \begin{array}{ll}
		\displaystyle
    \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}
    \Y{i-1,j}{m- \Kron_{a',s_i}}{a',b} &
    \text{Elif }S_i=-1 \\
    \displaystyle
    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
		 e^{\frac{-E_{(i,j),ab \to a'b'}^{\Omega,\beta}}{RT}}\cdot
    \Y{i-1,j+1}{m- \Kron_{a'b',s_is_j}}{a',b'} &
   	 \text{Elif }S_{i}=j \land S_{i+1}=j-1\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
  		 e^{\frac{-E_{(i,k),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
		 \cdot\Y{i-1,k+1}{m- \Kron_{a'b',s_is_k} - m'}{a',b'}
     \cdot\Z{j,k-1}{m'}{b,b'} &
		 \text{Elif }S_{i}=k \geq j\\
		 \displaystyle
		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_ks_i}\le m}}
		 \sum_{m'=0}^{m-\Kron_{a'b',s_ks_i}}
   	 e^{\frac{-E_{(k,i),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
		 \cdot\Y{k-1,j}{m- \Kron_{a'b',s_ks_i} - m'}{a',b}
     \cdot\Z{k+1,i-1}{m'}{a',b'} &
		 \text{Elif }-1 < S_{i}=k < i\\
		 0 & \text{Otherwise}
  \end{array}\right.
\label{eq:Y_rec}
\end{equation}
The five cases can be broked down as follows.
\begin{description}
\item[$S_i=-1$:] If the nucleotide at position $i$ is not paired, then the value is the same
as if we decrease the lower interval bound by $1$ (i.e. $i-1$), and consider all possible
nucleotides $a'$ at position $i$, correcting the number of mutants
in function of $\Kron_{a',s_i}$.
\item[$S_{i}=j$ and $S_{i+1}=j-1$:] If nucleotide $i$ is paired with $j$ and nucleotide $i+1$ is
paired with $j-11$, we are in the only case were stacked base pairs can occur. We thus add
the energy of the stacking and of the isostericity of the base pair $(i,j)$. What is left
to compute is the \emph{outside} value for the interval $[i-1,j+1]$ over all possible nucleotides 
$a',b'\in B^2$ at positions $i$ and $j$ respectively.
\item[$S_{i}=k \geq j$:]If nucleotide $i$ is paired with position $k\geq j$, 
and is not stacked inside, the 
only term contributing directly to the energy is the isostericity of the base pair $(i,k)$. 
We can then consider the outside interval $[i-1,k+1]$ by multiplying it by the the \emph{inside}
value of the newly included interval (i.e. $[j,k-1]$), for 
all possible values $a',b'\in B^2$ for nucleotides at positions $i$ and $k$ respectively.
\item[$-1<S_{i}<i$:]As above but if position $i$ is to pairing with a lower value.
\item[Else:] In all other cases, we are in a derivation of the SCFG that does not correspond to the 
secondary structure $S$, and we return $0$.

\end{description}

\subsubsection{Combining Inside and Outside values into point-wise mutations probabilities}
By construction, the partition function over all sequences at exactly $m$ mutations of $s$ can 
be described in function of the \emph{inside} term as $\Z{0,n-1}{m}{X,X}$,
 for any nucleotide $X\in B$ or
in function of the \emph{outside} term, for any unpaired position $k$:
$$
	\Z{0,n-1}{m}{X,X}
	\equiv
	\sum_{\substack{a\in \B,\\ \Kron_{a,s[k]}\le m}}	
	\Y{k-1,k+1}{m-\Kron_{a,s[k]}}{a,a}
$$

We are now left to compute the probability that a given position is a given nucleotide.
We leverage the \emph{Inside-Outside} construction to immediately obtain the following $3$ cases.
Given $i\in[0,n-1],x\in B$, and $M\geq 0$ a bound on the number of allowed mutations. 
\begin{align}
	\mathbb{P}(s_i = x\mid s,\Omega, S,M) &:= \frac{\mathcal{W}(i,x,s,\Omega,S,M)}{\sum_{m=0}^{M}\Z{0,n-1}{m}{X,X}}\label{eq:normalize}\\ 
\mathcal{W}(i,x,s,\Omega,S,M)&=
 \left\{
	\begin{array}{ll}
			\sum_{m=0}^{M}
			\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}
		&\text{If }S_i = -1\\
			\displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{xb,s_is_k\leq m}}}
			\sum_{m'=0}^{m-\Kron_{xb,s_is_k}}
     	 e^{\frac{-E_{(i,k),\varnothing\to xb}^{\Omega,\beta} }{RT}}
			\cdot\Y{i-1,k+1}{m-\Kron_{xb,s_is_k-m'}}{x,b}
			\cdot\Z{i+1,k-1}{m'}{x,b}
		&\text{If }S_i=k>i\\
    \displaystyle
			\sum_{m=0}^{M}
			\sum_{\substack{b\in B\\\Kron_{bx,s_ks_i\leq m}}}
			\sum_{m'=0}^{m-\Kron_{bx,s_ks_i}}
     	 e^{\frac{-E^{\Omega, \beta}_{(k,i),\varnothing\to bx}}{RT}}
			\cdot\Y{k-1,i+1}{m-\Kron_{bx,s_ks_i-m'}}{b,x}
			\cdot\Z{k+1,i-1}{m'}{b,x}
		&\text{If }S_i=k<i
	\end{array}\right.\label{eq:combine}
\end{align}

In every case, the denominator is the sum of the partitions function of exactly $m$ mutations, 
for $m$ smaller or equal to our target $M$. The numerators are divided in the following three cases.
\begin{description}
\item[$S_i=-1$:] If the nucleotide at position $i$ is not paired, we are concerned by the weights
over all sequences which have at position $i$ nucleotide $x$, which is exactly the sum
of the values of $\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}$, for all $m$ between $0$ and $M$.
\item[$S_i=k>i$:] Since we need to respect the derivation of the secondary structure $S$, if 
position $i$ is paired, we must consider the two partition functions. The \emph{outside} of the 
base pair, and the \emph{inside}, for all possible values for the nucleotide at position $k$, and
all possible distribution of the mutant positions between the inside and outside of the base pair. We also add the term of isostericity for this specific base pair.
\item[$S_i=k<i$:] Same as above, but with position $i$ pairing with a lower position.
\end{description}

\subsection{Complexity considerations}
Equations~\eqref{eq:Z_rec} and~\eqref{eq:Y_rec} can be computed using dynamic programming. Namely, the $\mathcal{Z}^{*}_{*}$ and $\mathcal{Y}^{*}_{*}$ terms are computed starting from smaller values of $m$ and interval lengths, memorizing the results as they become available to ensure constant-time access during later stages of the computation. Furthermore, energy terms $E(\cdot)$ can be accessed in constant time thanks to a simple precomputation (not described)  of the isostericity contributions in $\Theta(n\cdot|\Omega|)$. Computing any given term therefore requires $\Theta(m)$ operations.

>>>>

%
%\subsection{Probabilistic model}
%Let $\Omega$ be an gap-free RNA alignment sequence, $S$ its associated secondary structure, 
%then any sequence $s$ is assigned a probability proportional to its Boltzmann factor
%\begin{align*}
%  \mathcal{B}(s) &= e^\frac{-\PE{s}}{RT}, &&\text{with}&\PE{s}&:=\alpha\cdot\ES(s,S)+(1-\alpha)\cdot\EI(s,S,\Omega),
%\end{align*}
%where $R$ is the Boltzmann constant, $T$ the temperature in Kelvin, $\ES(s)$ and $\EI(s,S,\Omega)$ 
%are the free-energy and isostericity contributions respectively (further described below), and $\alpha\in[0,1]$ is an arbitrary parameter that sets the relative weight for both contributions.
%
%\subsubsection{Energy contribution}
%The free-energy contribution in our pseudo-energy model corresponds to an additive stacking-pairs model, taking values from the Turner 2004 model retrieved from the NNDB~\cite{Turner2010}. Given a candidate sequence $s$ for a secondary structure $S$, the free-energy of $S$ on $s$ is given by
%\begin{align*}
%  \ES(s,S) = \sum_{\substack{(i,j)\to (i',j')\in S\\ \text{stacking pairs}}}\ES^{\beta}_{s_is_j\to s_{i'}s_{j'}} 
%\end{align*}
%where $\ES^{\beta}_{ab\to a'b'}$ is set to $0$ if $ab=\varnothing$ (no base-pair to stack onto), the tabulated free-energy of stacking pairs $(ab)/(a'b')$ in the Turner model if available, or $\beta\in[0,\infty]$ for non-Watson-Crick/Wobble entries (i.e. not in $\{\Gb\Ub,\Ub\Gb,\Cb\Gb,\Gb\Cb, \Ab\Ub\text{ or }\Ub\Ab\}$). This latter parameter allows to choose whether to simply penalize invalid base pairs, or forbid them altogether ($\beta = \infty$).
%The loss of precision due to this simplification of the Turner model remains reasonable since the targeted secondary structure is fixed 
%(e.g. multiloops do not account for base-specific contributions). Furthermore, it greatly eases the design and implementation of dynamic-programming equations. 
%\subsubsection{Isostericity contribution}
%The concept of isostericity score~\cite{Stombaugh2009} is based on the geometric discrepancy (superimposability) of two base-pairs, using individual additive contributions computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Let $s$ be a candidate sequence for a secondary structure $S$, given in the context of a gap-free RNA alignment $\Omega$,  we define the contribution of the isostericity to the pseudo-energy as
%\begin{align*}
%  \ES(s,S,\Omega) &= \sum_{\substack{(i,j)\in S\\ \text{pairs}}}\EI^{\Omega}_{(i,j),s_i s_j}, & \text{where}&& 	\EI^{\Omega}_{(i,j),ab}:=
%	\frac{
%		\sum_{s'\in\Omega}
%			\text{\ISO}((s_i',s_j'),(a,b))}
%%-		\left(			\text{\ISO}((s_i',s_j'),(s_i,s_j))				\right)	
%{		
%		|\Omega|
%	}
%\end{align*}
%is the average isostericity of a base-pair in the candidate sequence, compared with the reference alignment.
%The $\ISO$ function uses the {Watson-Crick/Watson-Crick} cis isostericity matrix computed by Stombaugh~\emph{et al}~\cite{Stombaugh2009}. Isostericity scores range between $0$ and $9.7$, with $0$ being assigned to a perfect isostericity, and a penalty of $10$ is used for missing entries.
%The isostericity contribution will favor exponentially sequences that are likely to adopt a similar local conformation as the sequences contained in the alignment.
%
%\subsection{Mutational profile of sequences}
%
%
%Let $s$ be an RNA sequence, $S$ a reference structure, and $m\geq 0$ a desired number of mutations. 
%We are interested in  the probability that a given position contains a specific nucleotide, over all sequences having at most $m$ mutations from $s$
%%  under the SCFG derivation $S$ 
%(formally $\mathbb{P}(s_i = x\mid s,\Omega, S,m)$). 
%We define a variant of the
% \emph{Inside-Outside algorithm}~\cite{Lari1990}, allowing us to obtain the
% desired probability,  the two  functions
%$\mathcal{Z}_*^*$ and $\mathcal{Y}_*^*$. 
%
%The former, defined in Equations~\eqref{eq:Z_in} and~\eqref{eq:Z_rec}, is analogous to the \emph{inside} algorithm.
%It is the partition function, i.e. the sum of Boltzmann factors, over all sequences within $[i,j]$, 
%knowing that position $i-1$ is composed of nucleotide $a$ (resp. $j+1$ is $b$), within 
%$m$ mutations of $s$. The latter, defined by Equations~\eqref{eq:Y_in} and~\eqref{eq:Y_rec},
% computes the \emph{outside} algorithm, i.e.  
%the partition function over sequences within $m$ mutations of $s$, restricted to two intervals $[0,i]\cup[j,n-1]$, and 
%knowing  that position $i+1$ is composed of $a$ (resp. $j-1$ is $b$). A suitable combination of these terms, given in Equation~\eqref{eq:combine}, gives the total weight, and in turn the probability, of seeing a specific base at a given position.
%
%%Drawing a parallel with stochastic context-free grammars (SCFG), for which the inside-outside algorithm was introduced, the set of sequences can be seen as the language of words having length $n$, generated from a stochastic context-free grammar.
%%These rules would limit the number of mutations (i.e. )
%% $S$ can be considered as constraining weighing on the shape of parse trees the derivations in a classic 
%% of the {SCFG} generating  all  secondary structures of length  $|s|$. 
%
%
%
%
%
%\subsubsection{Definitions}
%Let $B:=\left\{\Ab,\Cb,\Gb,\Ub\right\}$ be the set of nucleotides.
%Given $s\in B^n$ an RNA sequence, let $s_i$ be the nucleotide at position $i$. Let $\Omega$ be a set of un-gapped RNA sequences of
%length $n$, and $S$ a secondary structure without pseudoknots. 
%Formally, if $(i,j)$ and $(k,l)$ are base pairs in $S$, there is no overlapping extremities
% $\{i,j\}\cap \{k,l\}=\varnothing$ and either the intersection is empty 
% ($[i,j]\cap[k,l]=\varnothing$) or one is included in the other ($[k,l]\subset[i,j]$ or 
% $[i,j]\subset[k,l])$. 
%
%Let us then remind a Hamming distance function $\delta: B^*\times B^* \to \mathbb{N}^+$, which takes two sequences $s'$ and $s''$ as input, $|s'|=|s''|$, and returns the number of differing positions.
%Finally, let us denote by $E_{(i,j),ab \to ab'}^{\Omega,\beta}$ the local contribution of a base-pair $(i,j)$ to the pseudo-energy, such that
%\begin{equation}
%  E_{(i,j),ab \to a'b'}^{\Omega,\beta}  = \alpha \cdot\ES^\beta_{a b \to a' b'}+(1-\alpha)\cdot\EI^{\Omega}_{(i,j),a'b'}.
%\end{equation}
%
%
%\subsubsection{Inside}
%\begin{figure}
%\resizebox{\textwidth}{!}{\input{FigDPInside.pgf}}
%\caption{Principle of the inside computation (aka partition function). Any sequence with $m$ mutations over an interval $[i,j]$ 
%can be decomposed as a sequence over $[i+1,j]$ preceded by a, possibly mutated, base at $i$ 
%(Unpaired case), a sequence over $[i+1,j-1]$ surrounded by some base-pair (Stacking-pair case), 
%or as two sequences over $[i+1,k-1]$ and $[k+1,j]$, completed by some base-pair (General base-pairing case). 
%In each case, one has to investigate any possible ways to distribute mutations between the different sequences 
%and locally instanciated bases.}
%\end{figure}
%
%The \emph{Inside} function $\Z{i,j}{m}{a,b}$ is the partition function, i.e. the sum of Boltzmann factors, over all sequences in the interval $[i,j]$, at distance $m$ of $s_{[i,j]}$, and having flanking nucleotides $a$ and $b$ (at positions $i-1$ and $j+1$ respectively). 
%Such terms can be defined by recurrence, for which the following initial conditions holds:
%
%\begin{equation}
%	\forall i \in [0,n-1]:\, \Z{i+1,i}{m}{a,b}=\left\{
%	\begin{array}{ll}
%		1 &\text{If } m = 0\\
%		0 &\text{Otherwise.}
%	\end{array}\right.
%\label{eq:Z_in}
%\end{equation}
%In other words, the set of sequences at distance $m$ of the empty sequence is either empty if $m>0$, or restricted to the empty sequence, having energy $0$, if $m=0$. Since the energetic terms only depend on base pairs, they are not involved in the initial conditions.

@@

%\subsubsection{Outside}	
%\begin{figure}
%\resizebox{\textwidth}{!}{\input{FigDPOutside.pgf}}
%\caption{Principle of the outside computation. Note that the outside algorithm uses intermediate results from the inside algorithm, 
%therefore its efficient implementation requires an implementation of the inside computation.}
%\end{figure}
%The \emph{Outside} function, $\mathcal Y$, is the partition function considering only the 
%contributions of subsequences $[0,i]\cup[j,n-1]$ over the mutants of $s$ having exactly $m$ mutations between $[0,i]\cup[j,n-1]$ and whose nucleotide at position $i+1$ is $a$ 
%(resp. in position $j-1$ it is $b$).
%We define function $\Y{i,j}{m}{a,b}$ as a recurrence, and will use as initial conditions:
%\begin{equation}
%	\Y{-1,j}{m}{X,X}:=
%		\displaystyle
%	  \Z{j,n-1}{m}{X,X}
%\label{eq:Y_in}
%\end{equation}
%The recurrence, as shown below, will increase the interval $[i,j]$ by decreasing $i$ when
%it is not base paired. If it is with a position $k>j$, we increase $j$ to include it.
% Thus, when we need
%to evaluate an interval as $(-1,j)$, all stems between $(0,j)$ are taken into account and the
%structure between $(j,n-1)$ must be a set of independent stems. Therefore,
% all the outside energy between $[j,n-1]$ is
%equal to $\Z{j,n-1}{m}{X,X}$, for any $X\in B$. The recursion itself is as follows.
%\begin{equation}
%	\Y{i,j}{m}{a,b} = \left\{
%  \begin{array}{ll}
%		\displaystyle
%    \sum_{\substack{a'\in \B,\\ \Kron_{a',s_i}\le m}}
%    \Y{i-1,j}{m- \Kron_{a',s_i}}{a',b} &
%    \text{Elif }S_i=-1 \\
%    \displaystyle
%    \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_j}\le m}}
%		 e^{\frac{-E_{(i,j),ab \to a'b'}^{\Omega,\beta}}{RT}}\cdot
%    \Y{i-1,j+1}{m- \Kron_{a'b',s_is_j}}{a',b'} &
%   	 \text{Elif }S_{i}=j \land S_{i+1}=j-1\\
%		 \displaystyle
%		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_is_k}\le m}}
%		 \sum_{m'=0}^{m-\Kron_{a'b',s_is_k}}
%  		 e^{\frac{-E_{(i,k),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
%		 \cdot\Y{i-1,k+1}{m- \Kron_{a'b',s_is_k} - m'}{a',b'}
%     \cdot\Z{j,k-1}{m'}{b,b'} &
%		 \text{Elif }S_{i}=k \geq j\\
%		 \displaystyle
%		 \sum_{\substack{a'b'\in \B^2,\\ \Kron_{a'b',s_ks_i}\le m}}
%		 \sum_{m'=0}^{m-\Kron_{a'b',s_ks_i}}
%   	 e^{\frac{-E_{(k,i),\varnothing\to a'b'}^{\Omega,\beta}}{RT}}
%		 \cdot\Y{k-1,j}{m- \Kron_{a'b',s_ks_i} - m'}{a',b}
%     \cdot\Z{k+1,i-1}{m'}{a',b'} &
%		 \text{Elif }-1 < S_{i}=k < i\\
%		 0 & \text{Otherwise}
%  \end{array}\right.
%\label{eq:Y_rec}
%\end{equation}
%The five cases can be broked down as follows.
%\begin{description}
%\item[$S_i=-1$:] If the nucleotide at position $i$ is not paired, then the value is the same
%as if we decrease the lower interval bound by $1$ (i.e. $i-1$), and consider all possible
%nucleotides $a'$ at position $i$, correcting the number of mutants
%in function of $\Kron_{a',s_i}$.
%\item[$S_{i}=j$ and $S_{i+1}=j-1$:] If nucleotide $i$ is paired with $j$ and nucleotide $i+1$ is
%paired with $j-11$, we are in the only case were stacked base pairs can occur. We thus add
%the energy of the stacking and of the isostericity of the base pair $(i,j)$. What is left
%to compute is the \emph{outside} value for the interval $[i-1,j+1]$ over all possible nucleotides 
%$a',b'\in B^2$ at positions $i$ and $j$ respectively.
%\item[$S_{i}=k \geq j$:]If nucleotide $i$ is paired with position $k\geq j$, 
%and is not stacked inside, the 
%only term contributing directly to the energy is the isostericity of the base pair $(i,k)$. 
%We can then consider the outside interval $[i-1,k+1]$ by multiplying it by the the \emph{inside}
%value of the newly included interval (i.e. $[j,k-1]$), for 
%all possible values $a',b'\in B^2$ for nucleotides at positions $i$ and $k$ respectively.
%\item[$-1<S_{i}<i$:]As above but if position $i$ is to pairing with a lower value.
%\item[Else:] In all other cases, we are in a derivation of the SCFG that does not correspond to the 
%secondary structure $S$, and we return $0$.
%
%
%\end{description}
%
%\subsubsection{Combining Inside and Outside values into point-wise mutations probabilities}
%By construction, the partition function over all sequences at exactly $m$ mutations of $s$ can 
%be described in function of the \emph{inside} term as $\Z{0,n-1}{m}{X,X}$,
% for any nucleotide $X\in B$ or
%in function of the \emph{outside} term, for any unpaired position $k$:
%$$
%	\Z{0,n-1}{m}{X,X}
%	\equiv
%	\sum_{\substack{a\in \B,\\ \Kron_{a,s[k]}\le m}}	
%	\Y{k-1,k+1}{m-\Kron_{a,s[k]}}{a,a}
%$$
%
%We are now left to compute the probability that a given position is a given nucleotide.
%We leverage the \emph{Inside-Outside} construction to immediately obtain the following $3$ cases.
%Given $i\in[0,n-1],x\in B$, and $M\geq 0$ a bound on the number of allowed mutations. 
%\begin{align}
%	\mathbb{P}(s_i = x\mid s,\Omega, S,M) &:= \frac{\mathcal{W}(i,x,s,\Omega,S,M)}{\sum_{m=0}^{M}\Z{0,n-1}{m}{X,X}}\label{eq:normalize}\\ 
%\mathcal{W}(i,x,s,\Omega,S,M)&=
% \left\{
%	\begin{array}{ll}
%			\sum_{m=0}^{M}
%			\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}
%		&\text{If }S_i = -1\\
%			\displaystyle
%			\sum_{m=0}^{M}
%			\sum_{\substack{b\in B\\\Kron_{xb,s_is_k\leq m}}}
%			\sum_{m'=0}^{m-\Kron_{xb,s_is_k}}
%     	 e^{\frac{-E_{(i,k),\varnothing\to xb}^{\Omega,\beta} }{RT}}
%			\cdot\Y{i-1,k+1}{m-\Kron_{xb,s_is_k-m'}}{x,b}
%			\cdot\Z{i+1,k-1}{m'}{x,b}
%		&\text{If }S_i=k>i\\
%    \displaystyle
%			\sum_{m=0}^{M}
%			\sum_{\substack{b\in B\\\Kron_{bx,s_ks_i\leq m}}}
%			\sum_{m'=0}^{m-\Kron_{bx,s_ks_i}}
%     	 e^{\frac{-E^{\Omega, \beta}_{(k,i),\varnothing\to bx}}{RT}}
%			\cdot\Y{k-1,i+1}{m-\Kron_{bx,s_ks_i-m'}}{b,x}
%			\cdot\Z{k+1,i-1}{m'}{b,x}
%		&\text{If }S_i=k<i
%	\end{array}\right.\label{eq:combine}
%\end{align}
%
%In every case, the denominator is the sum of the partitions function of exactly $m$ mutations, 
%for $m$ smaller or equal to our target $M$. The numerators are divided in the following three cases.
%\begin{description}
%\item[$S_i=-1$:] If the nucleotide at position $i$ is not paired, we are concerned by the weights
%over all sequences which have at position $i$ nucleotide $x$, which is exactly the sum
%of the values of $\Y{i-1,i+1}{m-\Kron_{x,s_i}}{x,x}$, for all $m$ between $0$ and $M$.
%\item[$S_i=k>i$:] Since we need to respect the derivation of the secondary structure $S$, if 
%position $i$ is paired, we must consider the two partition functions. The \emph{outside} of the 
%base pair, and the \emph{inside}, for all possible values for the nucleotide at position $k$, and
%all possible distribution of the mutant positions between the inside and outside of the base pair. We also add the term of isostericity for this specific base pair.
%\item[$S_i=k<i$:] Same as above, but with position $i$ pairing with a lower position.
%\end{description}
%
%\subsection{Complexity considerations}
%Equations~\eqref{eq:Z_rec} and~\eqref{eq:Y_rec} can be computed using dynamic programming. Namely, the $\mathcal{Z}^{*}_{*}$ and $\mathcal{Y}^{*}_{*}$ terms are computed starting from smaller values of $m$ and interval lengths, memorizing the results as they become available to ensure constant-time access during later stages of the computation. Furthermore, energy terms $E(\cdot)$ can be accessed in constant time thanks to a simple precomputation (not described)  of the isostericity contributions in $\Theta(n\cdot|\Omega|)$. Computing any given term therefore requires $\Theta(m)$ operations.
%
%In principle, $\Theta(m\cdot n^2)$ terms, identified by $(m,i,j)$ triplets, should be computed.
%However, a close inspection of the recurrences reveals that the computation can be safely restricted to a subset of intervals $(i,j)$.
